---
output: html_document
---
# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
If the file activity.csv is not found in the current directory, it is extracted from the activity.zip file and is loaded into "activity_data".

```{r get the raw data, echo=T}
if (!file.exists("./activity.csv")) {
  unzip("activity.zip", files = "activity.csv", overwrite = T)
}
activity_data <- read.csv("activity.csv")
```
```{r headers , echo=F}
headers <- names(activity_data)
```

By examining the activity_data we can see that it has the following columns: `r headers`.

## What is mean total number of steps taken per day?
We will need to first calculate the total number of steps per day before ploting the histogram.

```{r histogram of steps per day, echo=T}
steps.date <- aggregate(steps ~ date, data = activity_data, FUN = sum)
hist(steps.date$steps, breaks = 10, xlab = "Daily Number of Steps", main = "Histogram of steps per day")
```

```{r mean total steps, echo=T}
mean_total_steps <- mean(steps.date$steps, na.rm = T)
median_total_steps <- median(steps.date$steps, na.rm = T)
```

The mean total number of steps taken per day is `r mean_total_steps` and the median is `r median_total_steps`.

## What is the average daily activity pattern?
```{r daily activity pattern, echo=T}
convert_to_time <- function(x) {
  paste0(as.integer(x/100),":", round((x/100 - as.integer(x/100)) * 100))
}
z <- tapply(na.omit(activity_data)$steps, FUN = function(x) {mean(x)}, INDEX = as.factor(na.omit(activity_data)$interval))
max_interval <- names(which(z == max(z)))
max_avg_steps <- z[(which(z == max(z)))[1]]
plot(z, xlab = "Interval", ylab = "Average number of steps", type ="l", x = levels(factor(na.omit(activity_data)$interval)))
abline(v = as.numeric(names(which(z == max(z)))), col=3)
```

The 5-minute interval in which on average across all the days contains the most number of steps is `r max_interval`.

## Imputing missing values

```{r how many rows with missing values, echo=T}
rows_missing_values <- nrow(activity_data) - nrow(na.omit(activity_data))
```

The activity data  has `r rows_missing_values` rows with missing values. Below, the NA's are replaced with the mean number of steps for the 5-minute interval.

```{r replace the missing values with mean value and show histogram, echo=T}
f_activity_data <- activity_data
for (j in 1:nrow(f_activity_data)){
  f_activity_data$steps[j] <- ifelse (is.na(f_activity_data$steps[j]), z[as.factor(f_activity_data$interval)][j], f_activity_data$steps[j])
  }
f_steps.date <- aggregate(steps ~ date, data = f_activity_data, FUN = sum)
hist(f_steps.date$steps, breaks = 10, xlab = "Daily Number of Steps", main = "Histogram of steps per day\nafter replacing NA\'s with the\naverage number of steps for that interval")
```

And now recalculate the mean and median.

```{r adjusted mean total steps, echo=T}
f_mean_total_steps <- mean(f_steps.date$steps)
f_median_total_steps <- median(f_steps.date$steps)
```

Once the NA's are replaced with average number of steps for that interval, the new average of total number of steps taken per day is `r f_mean_total_steps` (compared to `r mean_total_steps` before) and the new median is `r f_median_total_steps` (compared to `r median_total_steps` before).

The values don't change ina a meaningful way as expected, since the NA's were replaced by the averages.

## Are there differences in activity patterns between weekdays and weekends?

We will first separate the weekdays and the weekends and then create two plots, one for the weekends and one for the weekdays.

```{r are there differences between weekdays and weekends}
weekdays <- function(x) {
  format(as.Date(x), "%a") %in% c("Mon", "Tue", "Wed", "Thu", "Fri")
}
w_activity_data <- cbind(activity_data, weekdays(activity_data$date))
names(w_activity_data) <- c(headers, "weekday")
z_weekday <- tapply(subset(na.omit(w_activity_data), weekday)$steps, mean, INDEX = as.factor(subset(na.omit(w_activity_data), weekday)$interval))
z_weekend <- tapply(subset(na.omit(w_activity_data), !weekday)$steps, mean, INDEX = as.factor(subset(na.omit(w_activity_data), !weekday)$interval))
par(mfrow = c(2, 1), oma = c(2, 2, 0, 0), mar=c(3,4,2,2), mgp = c(2, 1, 0), xpd = NA)
plot(z_weekend, xlab = NA, main="Weekend", ylab = "Average Daily Steps", type ="l", x = levels(factor(na.omit(w_activity_data)$interval)))
axis(side = 1, labels =F)
axis(side = 2, labels =F)
plot(z_weekday, xlab = "Interval", main="Weekday", ylab = "Average Daily Steps", type ="l", x = levels(factor(na.omit(w_activity_data)$interval)))
axis(side = 1, labels =T)
axis(side = 2, labels =T)
```

This plot indicates that the pattern for the weekdays and weekends are different. The primary difference seems to be that most of the steps are taken in morning, prior to work hours during the week, while during the weekend it is more evenly spread out.